---
title: "MAAMOUL Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{maamoul_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MAAMOUL)
library(igraph)
library(tidyverse)
```

## 1. Preparing input data

MAAMOUL requires three main inputs (further described below):

1. A global metabolic network.  
2. P-values for enzymes (ECs) or gene families.     
3. P-values for metabolites. These p-values are derived from metabolomic data, indicating the association between each metabolite’s abundance and the phenotype of interest.       

In this tutorial, we load example datasets provided with the package, based on data from Franzosa et al., __Nature Microbiology__, 2019. We also describe how users can generate their own inputs.

### 1.1 Global metabolic network

The global metabolic network defines the relationships between metabolites and enzymes.
It can be derived from databases like KEGG or MetaCyc.    
The network should be represented as a table of edges connecting metabolite nodes to EC nodes, saved as a `.csv` file. The network should be bipartite, meaning that edges only connect nodes of different types (metabolites to ECs, not ECs to ECs or metabolites to metabolites).     

For this tutorial, we use a pre-compiled network (automatically loaded with the package):

```{r}
data(edges)
head(edges)
```

Note:   
* Each row in the table represents an edge  
* Only the first two columns are used; additional columns are ignored.

#### Building your own network

Users can construct a network from databases such as:
* KEGG   
* MetaCyc   
* Other metabolic pathway resources   

When building a custom network, we recommend:
* Removing very small connected components   
* Removing nodes with extremely high degree (hub nodes)   
  → These can dramatically influence network topology. Especially important for currency metabolites (e.g., water, ATP) that connect to many reactions and may not be informative for the specific biological question.   
* Filtering irrelevant ECs or metabolites based on prior knowledge   
  → e.g., ECs not linked to bacteria when studying the gut microbiome
  
#### Basic network diagnostics

To better understand your network, it is useful to inspect:
* Node degree distribution   
* Number and size of connected components   
* Total number of nodes and edges   

```{r fig.height = 2, fig.width = 4}
g <- graph_from_data_frame(edges, directed = FALSE)

# Degree distribution
deg <- degree(g)
ggplot(data.frame(deg), aes(x = deg)) +
  geom_histogram(color = 'black', fill = 'grey50', bins = 20) +
  ggtitle("Node degree distribution") +
  xlab('Node degree') +
  ylab('Count') +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))

# Connected components
comp <- components(g)
sprintf("Number of connected components: %i", comp$no)
sprintf("Largest component size: %i", max(comp$csize))
sprintf("Smallest component size: %i", min(comp$csize))

# Basic stats
sprintf("Number of EC nodes: %i", sum(grepl("^EC", V(g)$name)))
sprintf("Number of metabolite nodes: %i", sum(!grepl("^EC", V(g)$name)))
sprintf("Number of edges: %i", ecount(g))
```

### 1.2 EC (enzyme) p-values

The second input is a table of p-values for enzyme functions (ECs), saved in a `.tsv` file.    
These p-values are based on the analysis of metagenomic or metatranscriptomic data, reflecting the association between each EC’s abundance and the phenotype of interest (e.g., disease status). They are usually computed using differential abundance analysis methods (e.g., [MaAsLin3](https://www.nature.com/articles/s41592-025-02923-9), [ALDEx2](https://link.springer.com/article/10.1186/2049-2618-2-15), or [ANCOM](https://www.nature.com/articles/s41467-020-17041-7)).  

Load pre-computed EC p-values:

```{r}
data(ec_pvals)
ec_pvals %>% select(feature, pval) %>% head()
```

Notes:   
* EC identifiers must match those given in the global network table;   
* P-value may represent associations in both directions (enriched or depleted in the phenotype of interest), or in a single direction (using one-sided statistical tests). In the pre-computed example, p-values are two-sided, representing any association with the disease.   
* The table should include at least the following columns: `feature` → EC identifier (must match network node names), `pval` → statistical significance of association with the phenotype, before multiple testing correction. Additional columns (e.g., effect size, adjusted p-value) can be included but are not required.

### 1.3 Metabolite p-values

Similarly, MAAMOUL requires p-values for metabolites, derived from metabolomic data. 

```{r}
data(mtb_pvals)
mtb_pvals %>% select(feature, pval) %>% head()
```

Metabolite identifiers must match node names in the metabolic network.

### Inspect input data

It is useful to inspect how well the observed features overlap with the network. If most of the observed features are missing from the network, consider building a custom network that better captures the observed features.

```{r}
all_net_nodes <- unique(c(edges[[1]], edges[[2]]))

print("Loaded network information and feature p-values.")

sprintf(
  "%i of %i observed EC features (features for which a p-value has been computed) are also in the network.",
  sum(ec_pvals$feature %in% all_net_nodes),
  nrow(ec_pvals)
)

sprintf(
  "%i of %i observed metabolite features (features for which a p-value has been computed) are also in the network.",
  sum(mtb_pvals$feature %in% all_net_nodes),
  nrow(mtb_pvals)
)
```

